Exploring mathematical objects from custom-tailored mathematical universes

### 2019-01-09 paper-filmat.tex
to do
Some details on the relation to models of set theory

@@

* Andrej's realizability in the real world
* Cauchy vs. Dedekind numbers (physical quantities, ...)
* Syntactical vs. semantical interpretation
* phone call analogy
* detailed explanation of the pretty picture; analogy with "inner models"
  of set theory
* different intention with the alternative universes in comparison to set theory
* quick overview of the several aspects of toposes (maybe at the end,
  as an outlook?)
* enrichment of platonism debate (find better term for this!)
* uncovering further relations between objects
* allowing a switch of perspective
* applications in mathematical practice
* int. logic as common denominator
* arbitrariness of the "standard axioms"
* models of ZF yield toposes (models of ETCS), including a quick discussion of
  equivalence
* what giving up classical logic actually amounts to in practice
* examples in Eff
* examples in Eff(STM)
* examples in Sh(X)
* examples in Spec(A), "reifying all the individual localizations into a
  single coherent entity which can be reasoned about as if it were a single
  ring"
* importance of having an adapted language
* SDG (infinitesimals are well-studied in philosophy of mathematics)
* beautiful and intriguing fact: laws of logic apply to mathematical objects
  not only on the face, but also in different ways
* quick remark on the internal language being based on types instead of sets
* reference to Bohr topos approach
* Source https://plato.stanford.edu/entries/category-theory/ for relevant
  literatur
* the internal language as a generalized modal operator

Send to Georg.

>>>>

\section{Unsorted}

Toposes can be pictured as ``universes in which we can do mathematics in'' in
much the same way as models of set theory can be viewed in this way. In fact,
to any model~$M$ of a set theory such as~$\mathbf{ZF}$ or~$\mathbf{ZFC}$, there
is a topos~$\Set_M$ such that a statement holds in the internal language
of~$\Set_M$ if and only if it holds in~$M$.\footnote{The topos~$\Set_M$ can be
described as follows: Its objects are the elements of~$M$, that is the things
which~$M$ believes to be sets, and its morphisms are those things which~$M$
believes to be maps. The topos~$\Set_M$ validates the axioms of
$\mathbf{ETCS}$~\cite{XXX}, and for models which are not elementarily
equivalent, their associated toposes will not be equivalent.} This
includes~$V$, the trivial inner model consisting of all sets there are, in case
such a thing exists in one's chosen ontology; in this case the associated
topos~$\Set_V$ is the category of all sets, and a statement holds in~$\Set_V$
if and only if it's true in the ordinary mathematical sense.

Carrying the analogy further, just as using forcing and other techniques we can
construct new models of set theory from given ones, thereby exploring the
set-theoretic multiverse, we can construct new toposes from given toposes.
However, there are two important differences between the notion of mathematical
universe as provided by toposes and as provided by models of set theory, both
regarding the subject matter and the reasons for why we are interested in them.

Firstly, toposes are more general than models of set theory. By definition, a
model of~$\mathbf{ZFC}$ will always satisfy the axioms of~$\mathbf{ZFC}$; in
contrast, most toposes do not even validate the law of excluded middle, much
less so the axiom of choice (which implies the law of excluded middle in
presence of the remaining axioms of intuitionistic Zermelo--Fraenkel set
theory~\cite{XXX}). This failure of classicality is often for interesting
geometric or algebraic reasons instead of because of logical issues, as will be
detailed in Section~\ref{several-toposes}.

Secondly, there is a difference in motivation. The main philosophical reason
for studying models of set theory is to study which notions of sets are
coherent: Does the cardinality of the reals need to be the cardinal directly
succeeding~$\aleph_0$, the cardinality of the naturals? No, there are models of
set theory in which the continuum hypothesis fails. Do non-measurable sets of
reals need to exist? No, in models of~$\mathbf{ZF}+\mathbf{AD}$,
Zermelo--Fraenkel set theory plus the axiom of determinacy, it's a theorem that
every subset of~$\RR^n$ is Lebesgue-measurable. Can the axiom of choice be
added to the axioms of~$\mathbf{ZF}$ without causing inconsistency? Yes, if~$M$
is a model of~$\mathbf{ZF}$ then~$L^M$, the set of definable sets of~$M$, is a
model of~$\mathbf{ZFC}$.~\cite{XXX, maybe SEP?}

While toposes can be used for similar such purposes, and indeed have been,
especially to explore the various intuitionistic notions of sets, an important
aspect of topos theory is that toposes are used to explore the standard
mathematical universe: Truth in the \effective topos tells us what is
computable; truth in sheaf toposes tells us what's true locally, in a fashion
which depends continuously on parameters; the little Zariski topos of a general
commutative ring can be used to pretend that the given ring is a field; toposes
adapted to synthetic differential geometry can be used to rigorously work with
infinitesimals. All of these examples will be presented in more detail in
Section~\ref{several-toposes}.

In a certain precise way, toposes allow us to study the common objects of
mathematics from a different point of view -- one such view for every topos --
and it is a beautiful and intriguing fact that with the sole exception of the
law of excluded middle, the laws of logic apply to mathematical objects also
when viewed through the lens of a specific topos.

@@

* Andrej's realizability in the real world
* Cauchy vs. Dedekind numbers (physical quantities, ...)
* Syntactical vs. semantical interpretation
* phone call analogy
* detailed explanation of the pretty picture; analogy with "inner models"
  of set theory
* different intention with the alternative universes in comparison to set theory
* quick overview of the several aspects of toposes (maybe at the end,
  as an outlook?)
* enrichment of platonism debate (find better term for this!)
* uncovering further relations between objects
* allowing a switch of perspective
* applications in mathematical practice
* int. logic as common denominator
* arbitrariness of the "standard axioms"
* models of ZF yield toposes (models of ETCS), including a quick discussion of
  equivalence
* what giving up classical logic actually amounts to in practice
* examples in Spec(A), "reifying all the individual localizations into a
  single coherent entity which can be reasoned about as if it were a single
  ring"
* importance of having an adapted language
* SDG (infinitesimals are well-studied in philosophy of mathematics)
* quick remark on the internal language being based on types instead of sets
* reference to Bohr topos approach
* Source https://plato.stanford.edu/entries/category-theory/ for relevant
  literatur
* the internal language as a generalized modal operator

Send to Georg.
Also Giuseppe Cammarata.
